---
title: "Maize Production Analysis Across 47 Counties in Kenya (2022)"
description: "Statistical analysis of 2022 maize data: Rift Valley vs Non-Rift production, rainfall-yield relationship, and multiple regression modeling."
date: "2026-05-10"
categories: [R, Kenya, Agriculture, Regression]
---

## Overview

**Objective:** To analyze 2022 maize production across all 47 counties to answer three research questions for CAT 1.

**Dataset:** `kenya_maize_2022.xlsx` - 47 rows (counties), 6 columns Variables: `county`, `area_planted_ha`, `production_bags_90kg`, `yield_bags_per_ha`, `rainfall_mm`, `fertilizer_used_tons`

### Tools

`readxl` `ggplot2` `dplyr` `tidyr`

### Data Loading & Checking

```{r}
#| eval: false 
#| echo: true 
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Load dataset - USE RELATIVE PATH FOR WEBSITE

```{r}
#| eval: false 
kenya_maize_2022 <- read_excel("files/kenya_maize_2022-1.xlsx")
df <- kenya_maize_2022
```

# Check structure - should be 47 rows, 6 cols

```{r}
#| echo: false
#| eval: false
str(df)
summary(df)
sum(is.na(df))
```

## Question 1: Is Rift Valley Production Higher?

```{r}
#| echo: false
library(readxl) 
library(ggplot2) 
library(dplyr) 
library(tidyr) 
#loading dataset 
kenya_maize_2022 <- read_excel("files/kenya_maize_2022-1.xlsx")
df <- kenya_maize_2022 
rift_counties <- c("Trans Nzoia","Uasin Gishu","Nakuru","Narok","Nandi",
                   "Baringo","Elgeyo-Marakwet","Kericho","Bomet",
                   "Laikipia","Kajiado","West Pokot")

df$is_rift <- ifelse(df$county %in% rift_counties, "Rift","Non-Rift")
aggregate(production_bags_90kg ~ is_rift, data=df, mean)
t.test(production_bags_90kg ~ is_rift, data=df)

```

```{r}
#| echo: false
# Bar graph for all 47 counties
ggplot(df, aes(x=reorder(county, production_bags_90kg), 
               y=production_bags_90kg, fill=is_rift)) + 
  geom_col() +
  coord_flip() + 
  labs(title="Maize Production: Rift vs Non-Rift (All 47 Counties)", 
       x="County", y="Production (90kg bags)")

# Top 15 Producers
df_top <- df %>% arrange(desc(production_bags_90kg)) %>% head(15)
ggplot(df_top, aes(x=reorder(county, production_bags_90kg), 
                   y=production_bags_90kg, fill=is_rift)) +
  geom_col(width = 0.6) +
  coord_flip() +
  labs(title="Top 15 Producers: Rift vs Non-Rift", 
       x="County", y="Production (90kg bags)")
```

**Findings**: Rift Valley counties have significantly higher mean production (t-test p \< 0.05). Trans Nzoia, Uasin Gishu, Nakuru dominate.

## Question 2: Relationship Between Rainfall and Yield

```{r}
#| echo: false
#| message: false
#| warning: false
cor.test(df$rainfall_mm, df$yield_bags_per_ha)
model_yield <- lm(yield_bags_per_ha ~ rainfall_mm, data=df)
summary(model_yield)

ggplot(df, aes(x=rainfall_mm, y=yield_bags_per_ha)) +
  geom_point(size=2) +
  geom_smooth(method="lm", se=FALSE, color="red") +
  labs(title="Rainfall vs Yield - Strong Positive Relationship",
       x="Rainfall (mm)", y="Yield (bags per ha)")
```

**Result from analysis**:

Correlation r = 0.776 (strong positive)

Slope = 0.0169 (for every 1mm increase in rainfall, yield increases by 0.0169 bags/ha)

R² = 0.603 -\> Rainfall explains 60.3% of variation in yield.

This is a key finding for food security policy.

## Question 3: Multiple Regression Models

```{r}
#| echo: false
# Model 1: What drives total production?
model_prod <- lm(production_bags_90kg ~ area_planted_ha + rainfall_mm + 
                   fertilizer_used_tons, data=df)
summary(model_prod)

# Model 2: What drives yield per hectare?
model_yield_fert <- lm(yield_bags_per_ha ~ rainfall_mm + fertilizer_used_tons, 
                       data=df)
summary(model_yield_fert)
```

These models show the combined effect of area, rainfall, and fertilizer - critical for government subsidy planning.

**Conclusion & Recommendations**

Rift Valley remains Kenya's breadbasket - prioritize storage infrastructure there.

Rainfall is the biggest driver of yield (60% explained) - invest in irrigation for low-rainfall counties.

Fertilizer use should be analyzed further for yield gap closure.

**Files**

::: {.callout-note title="Download Project Files"}
- <a href="files/Maize_Production_CAT1.pdf" download>📄 Download Full Report (PDF)</a>
- <a href="files/kenya_maize_2022-1.xlsx" download>📊 Download Dataset (Excel - 47 Counties)</a>
:::
